---
title: "Sistemas y Señales Biomédicos"
description: "SYSB"
subtitle: "Ingeniería Biomédica"
lang: es
author: "Ph.D. Pablo Eduardo Caicedo Rodríguez"
date: "2025-01-20"
format:
  revealjs: 
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    code-copy: true
    fig-align: center
    self-contained: true
    theme: 
      - simple
      - ../../recursos/estilos/metropolis.scss
    slide-number: true
    preview-links: auto
    logo: ../../recursos/imagenes/generales/Escuela_Rosario_logo.png
    css: ../../recursos/estilos/styles_pres.scss
    footer: <https://pablocaicedor.github.io/>
    transition: fade
    progress: true
    scrollable: true

resources:
  - demo.pdf
---

```{r}
#| echo: false
#| eval: true
#| output: false
#| label: Loading R-Libraries
# install.packages(c("DiagrammeR", "reticulate", "kableExtra", "tidyverse", "knitr", "cowplot", "ggfx"))
library("DiagrammeR")
library("reticulate")
library("kableExtra")
library("tidyverse")
library("knitr")
library("cowplot")
library("ggfx")
knitr::opts_chunk$set(echo = FALSE)

def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
})
```

```{python}
# | echo: false
# | eval: true
# | output: false
# | label: Loading Python-Libraries

import numpy as np
import matplotlib.pyplot as plt

plt.rcParams.update(
    {
        "text.usetex": True,  # usar LaTeX real
        "font.family": "Fira Code",  # familia general
        "mathtext.fontset": "custom",  # fuente personalizada para fórmulas
        "mathtext.rm": "Fira Code",  # texto “roman”
        "mathtext.it": "Fira Code:italic",  # texto itálico
        "mathtext.bf": "Fira Code:bold",  # texto en negrita
        "text.latex.preamble": r"\usepackage{amsmath}",
    }
)

path_ecg = "../../data"

```

# Sistemas y Señales Biomedicos - SYSB

## Signal Classification

:::: {.columns}

::: {.column width="45%"}

```{r}
#| echo: false
#| eval: true
#| output: true
#| fig-align: center
#| label: Mapa Conceptual 01

mermaid("
graph TD
  A[Signal Classification]
  B[Time]
  C[Amplitude]
  D[Continous]
  E[Discrete]
  F[Continous]
  G[Discrete]
  A --> B
  A --> C
  B --> D
  B --> E
  C --> F
  C --> G
")


```

:::

::: {.column width="45%"}

```{python}
#| echo: false
#| eval: true
#| output: true
#| label: Graph 01

# Parameters
signal_frequency = 5  # Frequency of the signal in Hz
sampling_frequency = (
    100  # Sampling frequency in Hz (at least 2 * signal_frequency for Nyquist)
)
duration = 1  # Duration of the signal in seconds
quantization_levels = 8  # Number of quantization levels

# Generate the continuous signal
t_continuous = np.linspace(0, duration, 1000)  # High-resolution time vector
continuous_signal = np.sin(2 * np.pi * signal_frequency * t_continuous)

# Generate the sampled signal
t_sampled = np.arange(0, duration, 1 / sampling_frequency)  # Sampling instants
sampled_signal = np.sin(2 * np.pi * signal_frequency * t_sampled)

# Quantize the sampled signal
quantized_signal = np.round(sampled_signal * (quantization_levels / 2 - 1)) / (
    quantization_levels / 2 - 1
)

# Zero-order hold quantization: extend the quantized value to the next sampling time
t_zero_hold = np.repeat(t_sampled, 2)[1:]  # Repeat and shift time vector
quantized_zero_hold = np.repeat(quantized_signal, 2)[
    :-1
]  # Repeat quantized values for hold effect

# Create subplots
fig, axs = plt.subplots(3, 1, figsize=(12, 10), sharex=True)

# Plot continuous signal
axs[0].plot(
    t_continuous,
    continuous_signal,
    label="Continuous Signal",
    linewidth=2,
    color="blue",
)
axs[0].set_title("Continuous Signal", fontsize=14)
axs[0].set_ylabel("Amplitude", fontsize=12)
axs[0].grid(True)
axs[0].legend(fontsize=12)

# Plot sampled signal
axs[1].stem(
    t_sampled,
    sampled_signal,
    linefmt="r",
    markerfmt="ro",
    basefmt="gray",
    label="Sampled Signal",
)
axs[1].set_title("Sampled Signal", fontsize=14)
axs[1].set_ylabel("Amplitude", fontsize=12)
axs[1].grid(True)
axs[1].legend(fontsize=12)

# Plot zero-order hold quantized signal
axs[2].plot(
    t_zero_hold,
    quantized_zero_hold,
    label="Zero-Order Hold Quantized Signal",
    linestyle="-",
    color="green",
    linewidth=2,
)
axs[2].set_title("Zero-Order Hold Quantized Signal", fontsize=14)
axs[2].set_xlabel("Time (s)", fontsize=12)
axs[2].set_ylabel("Amplitude", fontsize=12)
axs[2].grid(True)
axs[2].legend(fontsize=12)

# Adjust layout
plt.tight_layout()
plt.show()
```

:::
::::

## Signal Classification -- Bounded

```{python}
#| echo: false
#| eval: true
#| output: true
#| fig-align: center
#| label: Graph 02

# Parameters
duration = 2  # Duration of the signals in seconds
t = np.linspace(0, duration, 1000)  # Time vector for continuous signals

# Bounded signal
bounded_signal = np.sin(2 * np.pi * 2 * t)  # Sinusoidal signal (bounded between -1 and 1)

# Unbounded signal
unbounded_signal = np.exp(t)  # Exponentially increasing signal (unbounded)

# Create subplots
fig, axs = plt.subplots(2, 1, figsize=(10, 8), sharex=True)

# Plot bounded signal
axs[0].plot(t, bounded_signal, label="Bounded Signal (Sine Wave)", color="blue", linewidth=2)
axs[0].set_title("Bounded Signal", fontsize=14)
axs[0].set_ylabel("Amplitude", fontsize=12)
axs[0].grid(True)
axs[0].legend(fontsize=12)

# Plot unbounded signal
axs[1].plot(t, unbounded_signal, label="Unbounded Signal (Exponential)", color="red", linewidth=2)
axs[1].set_title("Unbounded Signal", fontsize=14)
axs[1].set_xlabel("Time (s)", fontsize=12)
axs[1].set_ylabel("Amplitude", fontsize=12)
axs[1].grid(True)
axs[1].legend(fontsize=12)

# Adjust layout
plt.tight_layout()
plt.show()

```

## Signal Classification -- Compact Support

```{python}
#| echo: false
#| eval: true
#| output: true
#| fig-align: center
#| label: GRaph 03

# Parameters
duration = 2  # Total time in seconds
t = np.linspace(-1, 3, 1000)  # Time vector

# Compact support signal: parabola within a specific range (0 to 2)
compact_support_signal = np.where((t >= 0) & (t <= 2), 1 - (t - 1) ** 2, 0)  # Parabola only in [0, 2]

# Define the support range for highlighting
support_start, support_end = 0, 2

# Plot the signal
plt.figure(figsize=(10, 6))

# Plot the compact support signal
plt.plot(t, compact_support_signal, label="Compact Support Signal", color="blue", linewidth=2)

# Highlight the support in red
plt.axvspan(support_start, support_end, color="red", alpha=0.2, label="Support [0, 2]")

# Configure the plot
plt.title("Compact Support Signal", fontsize=14)
plt.xlabel("Time (s)", fontsize=12)
plt.ylabel("Amplitude", fontsize=12)
plt.legend(fontsize=12)
plt.grid(True)
plt.show()

```

## Signal Classification -- Causal

```{python}
#| echo: false
#| eval: true
#| output: true
#| fig-align: center
#| label: Graph 04

# Parameters
t = np.linspace(-3, 3, 1000)  # Time vector

# Causal signal: exponential decay starting at t = 0
causal_signal = np.where(t >= 0, np.exp(-t), 0)

# Non-causal signal: exponential decay starting before t = 0
non_causal_signal = np.where(t <= 0, np.exp(t), 0)

# Bilateral signal: exponential decay symmetric around t = 0
bilateral_signal = np.exp(-np.abs(t))

# Create subplots
fig, axs = plt.subplots(3, 1, figsize=(10, 10), sharex=True)

# Plot causal signal
axs[0].plot(t, causal_signal, label="Causal Signal", color="blue", linewidth=2)
axs[0].set_title("Causal Signal", fontsize=14)
axs[0].set_ylabel("Amplitude", fontsize=12)
axs[0].grid(True)
axs[0].legend(fontsize=12)

# Plot non-causal signal
axs[1].plot(t, non_causal_signal, label="Non-Causal Signal", color="red", linewidth=2)
axs[1].set_title("Non-Causal Signal", fontsize=14)
axs[1].set_ylabel("Amplitude", fontsize=12)
axs[1].grid(True)
axs[1].legend(fontsize=12)

# Plot bilateral signal
axs[2].plot(t, bilateral_signal, label="Bilateral Signal", color="green", linewidth=2)
axs[2].set_title("Bilateral Signal", fontsize=14)
axs[2].set_xlabel("Time (s)", fontsize=12)
axs[2].set_ylabel("Amplitude", fontsize=12)
axs[2].grid(True)
axs[2].legend(fontsize=12)

# Adjust layout
plt.tight_layout()
plt.show()
```

## Signal Classification - Even/Odd

:::: {.columns}

::: {.column width="45%"}

::: {.callout-note title="Even"}

$$f\left(t\right) = f\left(-t\right)$$
$$f\left[t\right] = f\left[-t\right]$$

:::

::: {.callout-note title="Odd"}

$$f\left(t\right) = -f\left(-t\right)$$
$$f\left[t\right] = -f\left[-t\right]$$

:::

:::

::: {.column width="45%"}

```{python}
#| echo: false
#| eval: true
#| output: true
#| fig-align: center
#| label: Graph 05

# Parameters
t = np.linspace(-5, 5, 1000)  # Time vector

# Even signal: cosine function (symmetric about the y-axis)
even_signal = np.cos(t)

# Uneven signal: sine function (anti-symmetric about the origin)
uneven_signal = np.sin(t)

# Create subplots
fig, axs = plt.subplots(2, 1, figsize=(10, 8), sharex=True)

# Plot even signal
axs[0].plot(t, even_signal, label="Even Signal (Cosine)", color="blue", linewidth=2)
axs[0].set_title("Even Signal", fontsize=14)
axs[0].set_ylabel("Amplitude", fontsize=12)
axs[0].grid(True)
axs[0].legend(fontsize=12)

# Plot uneven signal
axs[1].plot(t, uneven_signal, label="Uneven Signal (Sine)", color="red", linewidth=2)
axs[1].set_title("Uneven Signal", fontsize=14)
axs[1].set_xlabel("Time (s)", fontsize=12)
axs[1].set_ylabel("Amplitude", fontsize=12)
axs[1].grid(True)
axs[1].legend(fontsize=12)

# Adjust layout
plt.tight_layout()
plt.show()

```

:::
::::

## Signal Classification

::: {.callout-note title="Decomposition"}
All signal can be decomposed in two signals: one even, one odd.

$$x(t) = x_{even}(t) + x_{odd}(t)$$
:::

Where:

$$x_{even}(t) = \frac{x(t)+x(-t)}{2} $$
$$x_{odd}(t) = \frac{x(t)-x(-t)}{2} $$

## Example

::: {.callout-tip title="Example"}
Decompose the signal $x(t)=e^{t}$ into its even and odd parts
:::

## Example

$$x_{\text{even}}(t) = \frac{x(t) + x(-t)}{2}$$

$$x_{\text{odd}}(t) = \frac{x(t) - x(-t)}{2}$$

$$x(-t) = e^{-t}$$

$$x_{\text{even}}(t) = \frac{e^t + e^{-t}}{2} = \cosh(t)$$

$$x_{\text{odd}}(t) = \frac{e^t - e^{-t}}{2} = \sinh(t)$$

$$x(t) = x_{\text{even}}(t) + x_{\text{odd}}(t)$$

$$e^t = \cosh(t) + \sinh(t)$$
​


## Example

```{python}
# | echo: false
# | eval: true
# | output: true
# | fig-align: center
# | label: Graph 06

# Time vector
t = np.linspace(-3, 3, 500)

# Define the original function and its components
f_t = np.exp(t)  # Original function: e^t
f_even = (np.exp(t) + np.exp(-t)) / 2  # Even part: cosh(t)
f_odd = (np.exp(t) - np.exp(-t)) / 2  # Odd part: sinh(t)

# Create the subplots
fig, axs = plt.subplots(3, 1, figsize=(10, 12), sharex=True)

# Plot the original function
axs[0].plot(t, f_t, label=r"$f(t) = e^t$", color="blue", linewidth=2)
axs[0].set_title("Original Function", fontsize=14)
axs[0].set_ylabel("Amplitude", fontsize=12)
axs[0].legend(fontsize=12)
axs[0].grid(True)

# Plot the even part
axs[1].plot(
    t, f_even, label=r"$f_{\text{even}}(t) = \cosh(t)$", color="green", linewidth=2
)
axs[1].set_title("Even Part of the Function", fontsize=14)
axs[1].set_ylabel("Amplitude", fontsize=12)
axs[1].legend(fontsize=12)
axs[1].grid(True)

# Plot the odd part
axs[2].plot(t, f_odd, label=r"$f_{\text{odd}}(t) = \sinh(t)$", color="red", linewidth=2)
axs[2].set_title("Odd Part of the Function", fontsize=14)
axs[2].set_xlabel("Time (s)", fontsize=12)
axs[2].set_ylabel("Amplitude", fontsize=12)
axs[2].legend(fontsize=12)
axs[2].grid(True)

# Adjust layout
plt.tight_layout()
plt.show()

```

## Signal Transformations

### Types of Transformations
Signals can undergo two types of transformations:

1. **Independent variable transformations** (affect the time or input axis).
2. **Dependent variable transformations** (affect the amplitude or output axis).

---

## Independent Variable Transformations

### Time Scaling
- Definition: Changes the time scale of the signal.
\[
x(at), \quad a > 1 \text{ (compression)}, \quad 0 < a < 1 \text{ (expansion)}
\]
- Example: If \( x(t) = \sin(t) \), then \( x(2t) \) is compressed.

### Time Shifting
- Definition: Shifts the signal in time.
\[
x(t - t_0) \quad \text{(right shift if \( t_0 > 0 \), left shift if \( t_0 < 0 \))}
\]
- Example: \( x(t - 2) \) shifts the signal 2 units to the right.

### Time Reversal
- Definition: Flips the signal across the vertical axis.
\[
x(-t)
\]
- Example: If \( x(t) = t^2 \), then \( x(-t) = t^2 \) (even signal).

---

## Dependent Variable Transformations

### Amplitude Scaling
- Definition: Multiplies the amplitude by a scalar factor.
\[
a \cdot x(t), \quad a > 1 \text{ (amplification)}, \quad 0 < a < 1 \text{ (attenuation)}
\]
- Example: If \( x(t) = \sin(t) \), then \( 2x(t) \) doubles the amplitude.

### Amplitude Shifting
- Definition: Adds a constant value to the amplitude.
\[
x(t) + c
\]
- Example: If \( x(t) = \cos(t) \), then \( x(t) + 2 \) shifts the signal up by 2 units.

---

## Combined Transformations

### Example
Consider:
\[
y(t) = 2 \cdot x(3t - 1) + 1
\]
1. **Time compression**: \( x(3t) \) compresses the signal.
2. **Time shift**: \( x(3t - 1) \) shifts it to the right by 1 unit.
3. **Amplitude scaling**: \( 2 \cdot x(3t - 1) \) amplifies the signal.
4. **Amplitude shift**: \( +1 \) shifts it upward.

---

## Visualization Example in Python

```{python}
#| echo: false
#| eval: true
#| output: true
#| label: Graph 07


t = np.linspace(-2, 2, 500)
x_t = np.sin(2 * np.pi * t)

# Original Signal
y1 = x_t

# Transformation: 2 * sin(3t - 1) + 1
t_new = 3 * t - 1
y2 = 2 * np.sin(2 * np.pi * t_new) + 1

plt.figure(figsize=(10, 6))
plt.plot(t, y1, label="Original Signal: $x(t) = \sin(2\pi t)$", color="blue")
plt.plot(t, y2, label="Transformed Signal: $2\sin(2\pi(3t-1)) + 1$", color="red")
plt.legend(fontsize=12)
plt.grid()
plt.title("Signal Transformations")
plt.xlabel("Time (t)")
plt.ylabel("Amplitude")
plt.show()
```